---
title: "An Analysis of Firearm-Related Shooting Occurrences in Toronto"
author: 
  - Sean Eugene Chua
thanks: "Code and data are available at: https://github.com/jfhasj/opendata-toronto-analysis"
date: today
date-format: long
abstract: "Gun violence remains a significant public safety concern in Toronto that necessitates appropriate enforcement strategies to minimize. This paper analyzes police shooting occurrences in the city from 2014 to 2023, where data shows that incidents peaked during summer months and late hours of the day, and the number of shooting occurrences experienced constant decline following the introduction of stricter gun control measures. Findings also highlight the influence of seasonal factors and potential nighttime effects that may inform the necessary approaches to minimize firearm-related shootings. Understanding these dynamics is essential for not only reducing shootings but also crime as a whole most especially in urban environments."
format: 
  pdf:
    toc: true
    number-sections: true
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(opendatatoronto)
library(tidyverse)
library(knitr)
library(here)
```

# Introduction

The increasing prevalence of gun violence in densely-populated and urban settings has raised significant concerns among policymakers, law enforcement, and communities alike. Ontario experienced the largest increase in shooting occurrences from 2021 to 2022 [@statcan-perreault], and analyzing patterns and trends in shootings in this regard could become vital in understanding the broader ramifications of firearm-related incidents and crime in general. This paper aims to provide a comprehensive analysis of firearm-related shooting occurrences in Toronto from 2014 to 2023 using data from the Toronto Police Services available as a dataset on Open Data Toronto [@citeShootingOcc]. By examining this data, we seek to identify any potential arising trends that may assist law enforcement in developing appropriate strategies to mitigate the number of shootings in years to come.

Despite the existing literature on gun violence, there remains a lack of localized studies that specifically address the temporal aspect of shootings in Canada, more specifically Toronto since most studies have largely focused on national trends or on cities in the US where shootings often occur. This paper aims to bridge this gap by investigating the possible presence of similar trends within Toronto and how these could be explained within the Canadian context. This paper also aims to provide details of how patterns in shooting occurrences can be noticed over time, particularly in relation to seasonal variations and time-of-day effects.

Results show that the number of shooting incidents generally increased from 2014 to 2019, followed by a decrease from 2020 to 2023. This trend correlates with implemented nationwide laws aimed at reducing gun violence, including the ban on assault-style weapons after the 2020 Nova Scotia mass shooting. Furthermore, results indicate that summer and winter months exhibit the highest and lowest rates of shooting occurrences respectively. Additionally, we observed a trend based on the time of day, where shootings are more likely to occur during nighttime hours and less likely to occur during the day.

Understanding these patterns is crucial for developing targeted interventions to reduce gun violence in Toronto. This study has numerous implications, the most significant of which is its potential to provide valuable insights for creating safer environments. This paper is structured as follows: @sec-data delves into an overview of the data used and some results and findings on this data, and @sec-discussion presents discussions surrounding our results, along with limitations and recommendations for future research.

# Data {#sec-data}

## Overview

The dataset used in this analysis was taken from Open Data Toronto and downloaded using the `opendatatoronto` library (Gelfand 2022). It records every reported occurrence of shootings or firearm discharges (excluding suicides and police involvements). All the data analysis was done through R [@citeR], along with external R packages such as `tidyverse` [@citeTidy], `knitr` [@citeKnit], and `here` [@citeHere].

The dataset used is published by Toronto Police Services entitled "Shootings & Firearm Discharges", The data is refreshed quarterly and is accurate as of July 23, 2024. The raw data set contains information on the date and time — with varying granularities — of each reported shooting and the number of injuries and deaths, if any, among others. Moreover, it is worth noting that only one other dataset in the Open Data Toronto catalogue contains data about shootings; the dataset entitled "Police Annual Statistical Report - Shooting Occurrences" contains data only from 2014-2019 and is less granular with its data on shooting times and dates. Thus, it was not used in this analysis.

The variables used in this analysis include the following: `OCC_DATE`, `OCC_YEAR`, `OCC_MONTH`, and `OCC_TIME_RANGE`. A sample of the cleaned data is provided in @tbl-clean. Note that these columns from the original dataset were renamed for easier reference.

```{r}
#| label: tbl-clean
#| echo: false

data_path <- here("outputs", "data", "cleaned_data.csv")
cleaned_data <- read_csv(data_path, show_col_types = FALSE)

kable(head(cleaned_data, n = 5), format = "pipe", caption = "Sample of Cleaned Shooting Dataset")
```

Additional summary statistics are displayed in @tbl-summary below. We see that there were a total of 3,779 shooting occurrences from 2014-2023, giving a mean of about 378 shootings per year, equivalent to more than 1 occurrence per day. However, as we have seen from @tbl-clean, shootings do not actually happen every day (for example, the next shooting after January 2, 2014 occurs on January 8, 2014). In addition, consider the standard deviation of 90.7 shootings; this tells us that there is considerable variation in the number of shootings each year, as seen in @fig-linechart-yearly-total.

```{r}
#| label: tbl-summary
#| echo: false

# Read the cleaned data
data_path <- here("outputs", "data", "cleaned_data.csv")
cleaned_data <- read_csv(data_path, show_col_types = FALSE)

# Convert date column to Date type if not already done
cleaned_data$date <- as.Date(cleaned_data$date)

# Calculate total shootings
total_shootings <- nrow(cleaned_data)

# Calculate average shootings per year
average_shootings_per_year <- cleaned_data %>%
  group_by(year = year(date)) %>%
  summarize(total = n()) %>%
  summarize(average = mean(total))

# Calculate average shootings per month
total_years <- length(unique(year(cleaned_data$date)))
total_months <- total_years * 12
average_shootings_per_month <- total_shootings / total_months


# Calculate average shootings per day
average_shootings_per_day <- total_shootings / as.numeric(difftime(max(cleaned_data$date), min(cleaned_data$date), units = "days"))

# Calculate standard deviation of yearly shootings
std_dev_yearly_shootings <- cleaned_data %>%
  group_by(year = year(date)) %>%
  summarize(total = n()) %>%
  summarize(sd = sd(total))

# Create summary table
summary_table <- data.frame(
  Statistic = c("Total Shootings", 
                "Mean Shootings per Year", 
                "Mean Shootings per Month", 
                "Mean Shootings per Day",
                "Standard Deviation of Yearly Shootings"),
  Value = c(total_shootings, 
            round(average_shootings_per_year$average, 2), 
            round(average_shootings_per_month, 2), 
            round(average_shootings_per_day, 2),
            round(std_dev_yearly_shootings$sd, 2))
)

# Print the summary table using kable
kable(summary_table, format = "pipe", caption = "Summary Statistics for Shooting Occurrences in Toronto from 2014-2023")
```

\newpage

## Results and Findings

The succeeding graphs in this section display various relationships between shooting occurrences and the variables in @tbl-clean. @fig-linechart-yearly-total shows the number of shootings by year.

```{r}
#| label: fig-linechart-yearly-total
#| fig-cap: Total Number of Shooting Occurrences from 2014-2023
#| fig-pos: "H"
#| echo: false

data_path <- here("outputs", "data", "cleaned_data.csv")
cleaned_data <- read_csv(data_path, show_col_types = FALSE)

# Group the data by year and count the number of rows for each year
yearly_totals <- cleaned_data %>%
  group_by(year) %>%
  summarize(total = n())

# Create the line graph
ggplot(yearly_totals, aes(x = year, y = total)) +
  geom_line() +
  labs(x = "Year", y = "Number of Shootings") + 
  theme_minimal()
```

We can see that shootings increased in every year from 2014-2016, decreased in 2017, increased again in every year from 2017-2019, followed by decreases in every year from 2020-2023. \newpage It is worth asking how the number of shooting occurrences change throughout the seasons. @fig-stackedbar shows that among the seasons, summer has generally more shootings than any other; winter has the least.

```{r}
#| label: fig-stackedbar
#| fig-cap: Total Number of Shootings by Month
#| fig-pos: "H"
#| echo: false

data_path <- here("outputs", "data", "cleaned_data.csv")
cleaned_data <- read_csv(data_path, show_col_types = FALSE)
cleaned_data$date <- as.Date(cleaned_data$date)
cleaned_data$month_num <- month(cleaned_data$date)

# Group by month number and count total shootings
monthly_totals <- cleaned_data %>%
  group_by(month_num) %>%
  summarize(total = n(), .groups = 'drop')

# Add season column
monthly_totals$season <- case_when(
  monthly_totals$month_num %in% c(12, 1, 2, 3) ~ "Winter",
  monthly_totals$month_num %in% c(4, 5) ~ "Spring",
  monthly_totals$month_num %in% c(6, 7, 8, 9) ~ "Summer",
  monthly_totals$month_num %in% c(10, 11) ~ "Fall"
)

season_colors <- c("Winter" = "skyblue1", 
                   "Spring" = "darkgreen", 
                   "Summer" = "yellow3", 
                   "Fall" = "orange")

# Create the stacked bar chart
ggplot(monthly_totals, aes(x = month_num, y = total, fill = season)) +
  geom_bar(stat = "identity") +
   geom_text(aes(label = total, y = total), vjust = -0.25, color = "black", size = 3) +
  scale_x_continuous(breaks = 1:12, labels = 1:12) +
  scale_fill_manual(values = season_colors) +
  labs(x = "Month Number", y = "Total Shootings", 
       fill = "Season") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5),
        legend.position = "bottom")
```

Moreover, we see that at the peak of summer (around July and August) is when the highest monthly total number of shootings occur while the peak of winter (around February) is when the least number of shootings occur. \newpage

We investigate this more in @fig-season-monthly-avg where we show the average number of shootings per month by season.

```{r}
#| label: fig-season-monthly-avg
#| fig-cap: Average Number of Shootings per Month by Season
#| fig-pos: "H"
#| echo: false

data_path <- here("outputs", "data", "cleaned_data.csv")
cleaned_data <- read_csv(data_path, show_col_types = FALSE)

# Convert date to Date type and extract month number
cleaned_data$date <- as.Date(cleaned_data$date)
cleaned_data$month_num <- month(cleaned_data$date)

# Define seasons
winter_months <- c(12, 1, 2, 3)
spring_months <- c(4, 5)
summer_months <- c(6, 7, 8, 9)
fall_months <- c(10, 11)

# Add season column to cleaned_data
cleaned_data$season <- case_when(
  cleaned_data$month_num %in% winter_months ~ "Winter",
  cleaned_data$month_num %in% spring_months ~ "Spring",
  cleaned_data$month_num %in% summer_months ~ "Summer",
  cleaned_data$month_num %in% fall_months ~ "Fall"
)

# Calculate average shootings per month for each season
seasonal_averages <- cleaned_data %>%
  group_by(season) %>%
  summarize(total = n(), .groups = 'drop') %>%
  mutate(months_in_season = case_when(
    season == "Winter" ~ length(winter_months),
    season == "Spring" ~ length(spring_months),
    season == "Summer" ~ length(summer_months),
    season == "Fall" ~ length(fall_months)
  )) %>%
  mutate(average = total / months_in_season / n_distinct(cleaned_data$year))

seasonal_averages$season <- factor(seasonal_averages$season, 
                                   levels = c("Winter", "Spring", "Summer", "Fall"))
season_colors <- c("Winter" = "skyblue1", 
                   "Spring" = "darkgreen", 
                   "Summer" = "yellow3", 
                   "Fall" = "orange")

# Create the bar chart with averages displayed on top
ggplot(seasonal_averages, aes(x = season, y = average, fill = season)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = sprintf("%.1f", average)), vjust = -0.25, color = "black", size = 4) +
  scale_fill_manual(values = season_colors) +
  labs(x = "Season", y = "Average Number of Shootings per Month", fill = "Season") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5),
        legend.position = "none") +
  ylim(0, max(seasonal_averages$average) * 1.1) # Extend y-axis to make room for labels
```

We see that from 2014-2023, the number of shootings are highest over the summer months (with 36.7 occurrences per month on average) and at its lowest over the winter months (with 26.6 occurrences per month on average). \newpage @fig-timeofday-avg shows a great discrepancy between the number of shootings per year relative to the time of day.

```{r}
#| label: fig-timeofday-avg
#| fig-cap: Average Number of Shootings by Time-of-Day
#| fig-pos: "H"
#| echo: false
data_path <- here("outputs", "data", "cleaned_data.csv")
cleaned_data <- read_csv(data_path, show_col_types = FALSE)

# Calculate average shootings per time of day
time_of_day_averages <- cleaned_data %>%
  group_by(time_of_day) %>%
  summarize(total = n(), .groups = 'drop') %>%
  mutate(average = total / n_distinct(cleaned_data$year))

time_of_day_averages$time_of_day <- factor(time_of_day_averages$time_of_day, 
                                            levels = c("Morning", "Afternoon", "Evening", "Night"))

time_colors <- c("Morning" = "yellow3", 
                 "Afternoon" = "orange", 
                 "Evening" = "royalblue2", 
                 "Night" = "grey23")

# Create the bar chart with averages displayed on top
ggplot(time_of_day_averages, aes(x = time_of_day, y = average, fill = time_of_day)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = sprintf("%.1f", average)), vjust = -0.25, color = "black", size = 4) +
  scale_fill_manual(values = time_colors) +
  labs(x = "Time of Day", y = "Average Number of Shootings per Year", fill = "Time of Day") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none") +
  ylim(0, max(time_of_day_averages$average) * 1.1) # Extend y-axis to make room for labels
```

We see that shootings are about 5.2 to 5.5 more likely to occur in the evening (6pm to 11:59pm) or at night (12am to 5:59am) compared to morning hours (6am to 11:59am).

We discuss each of the above graphs further in @sec-discussion next.

# Discussion {#sec-discussion}

In @fig-linechart-yearly-total, we see a general increase in the number of shootings from 2014-2019 before successive decreases from 2020-2023. One reason for the decreasing trend from 2020-2023 could partially be attributed to the effects of the pandemic. National lockdowns forced people to stay inside their homes, minimizing face-to-face interaction and thus limiting opportunities for shootings.

However, a larger contributor to the sudden shift in the number of shootings between the aforementioned time periods is the Canadian government’s implementation of a ban on assault-style weapons as a result of the Nova Scotia mass shooting in 2020, which was the worst in Canadian history [@guardian-assault-ban]. This ban covers the buying, selling, importing, and owning of such weapons. Although handguns are most often used in shooting-related crimes such as homicides [@statcan-perreault], the ban on assault-style weapons still contributed to the decrease in the number of shootings. The Canadian government has implemented parts of Bill C-21, an initiative to strengthen gun control measures. One measure which has already been put in place is the aforementioned ban on assault-style weapons; another is the “National Handgun Freeze” which imposes “restrictions on the sale, purchase, transfer, or import of handguns by individuals (with limited exceptions)” according to [@bill-c21]. With this, we expect the number of shooting occurrences in 2024 to further decrease from previous years.

From @fig-stackedbar and @fig-season-monthly-avg above, we see that summer and winter months see the highest and lowest (number and) rate of shootings. This is consistent with what [@seasonal-influence] find in their study “Seasonal Influence on Mass Shootings.” In the years covered by their study (2013-2015), the researchers found that there was “a clear increase in the number of mass shootings as well as the numbers killed and injured during the summer”, which is consistent with our findings in Toronto. They also state that shootings are highest “in May and finishes around September – October” which we also see in @fig-stackedbar. Additional data also support this idea as found by [@statcan-moreau] where he showed the existence of a seasonal regularity of crime in Canada. He states that in 2021, crime levels are lowest during fall and winter and highest during summer. He also postulates that this phenomenon is due to higher and lower tendencies for socialization during summer and winter months respectively.

From @fig-timeofday-avg, we see that shootings occur most in the late hours of the day (at night and in the evening). This trend was also examined by [@uscrime-timeofday]; in analyzing some US cities, data showed that most shootings occurred at night and that such a nighttime effect exists. They also observed seasonality in crimes which was previously seen in Graphs 2 and 3 as well.

All this being said, there is substantial evidence of relationships between the number of shooting occurrences and the year where we saw a general increase and decrease from 2014-2019 and 2020-2023 respectively as shown in @fig-linechart-yearly-total. Moreover, there also exists seasonal trends in shooting occurrences from @fig-stackedbar and @fig-season-monthly-avg, where we saw that most shootings occur in the summer, and the least occur in the winter. Finally, the time of day also has an effect on the number of shootings, with a nighttime effect being prominent as shown in the study by [@uscrime-timeofday] and @fig-timeofday-avg.

The dataset used in this analysis does not take into account firearm-related shootings caused by police involvements and suicides. Although these are not technically crimes, their exclusion might affect how the use of firearms in Toronto and Canada is perceived and is something worth taking note of. An area of future research worth studying further is the correlation between the number of shooting occurrences and the socio-economic status of those committing the crime or the area in which the shootings occurred. This would provide a better idea on the “distribution” of shootings in different areas of Toronto, allowing law enforcement to focus strategic efforts in more affected areas.

# Appendix

## Data Cleaning

This dataset provided by the Toronto Police Services has incomplete data on location data before 2014, so we focus on 2014-2023. 2024 data is excluded, as it is currently incomplete. Columns not pertinent to this analysis were also removed, namely `X_id`, `EVENT_UNIQUE_ID`, `HOOD_158`, `NEIGHBOURHOOD_158`, `HOOD_140`, `NEIGHBOURHOOD_140`, and `geometry`.

# References
